---
title: "Real-data workflow: RPE1 Perturb-seq"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# Overview and scope

This tutorial runs CausalGRN on a prepared subset of the RPE1 Perturb-seq
dataset. It shows one complete application workflow:

1. infer a causal gene regulatory network (GRN) that includes a transcriptional
   program node;
2. fit the perturbation-effect prediction model;
3. predict expression changes after setting three target genes to specified
   expression levels; and
4. evaluate how accurately the predictions rank held-out differentially
   expressed genes (DEGs).

The companion
[manuscript-code repository](https://github.com/yub-hutch/CausalGRN-manuscript-code)
provides the broader cross-validation and method-comparison analyses.

The example takes approximately 15 minutes with 10 CPU cores and used about
18 GB of memory in our test.

```{r install, eval = FALSE}
# Run once if CausalGRN is not installed.
devtools::install_github("yub-hutch/CausalGRN")
```

```{r packages}
library(CausalGRN)
library(dplyr)
library(ggplot2)
library(igraph)
```

# Prepare the RPE1 inputs

## Data sources and preprocessing

The RPE1 data and the K562 source-context summaries originate from the
[processed Perturb-seq data of Replogle and
Weissman](https://doi.org/10.25452/figshare.plus.20029387). The HCT116
source-context summaries originate from
[X-Atlas/Orion](https://doi.org/10.25452/figshare.plus.29190726.v1).

The deposited tutorial file provides processed inputs ready for CausalGRN. The
Replogle release normalizes expression within experimental gem groups to
address batch effects. Our subsequent preparation filters cells and genes and
regresses dataset-specific technical covariates from normalized expression. In
another dataset, the appropriate covariates may differ; the general requirement
is to remove known technical and confounding effects, including batch effects,
before CausalGRN analysis.

The
[manuscript-code repository](https://github.com/yub-hutch/CausalGRN-manuscript-code)
documents the dataset-specific regression models and complete processing steps.
This tutorial starts from their prepared outputs and focuses on the executable
CausalGRN workflow.

## Gene-program input

RPE1 contains a dominant perturbation-response program. The companion
preprocessing workflow identifies this program and defines its member genes.
The deposited tutorial data provide the resulting program node and
`program_genes` input.

For a new dataset, users should inspect the perturbation responses for
influential programs. When a program is present, define its member genes and
include it in the network with the gene-program functions used below. When no
influential program is present, use `infer_skeleton()`, `infer_causalgrn()`,
`fit_expression_model()`, and `predict_perturbation_effect()` as shown in the
quick-start tutorial.

## Reference source effects

When reference Perturb-seq datasets are available, first score their cells
using the same program definition as the target dataset. For source dataset
\(s\) and perturbation \(g\), summarize the source effect as

\[
\Delta_{s,g} =
\operatorname{mean}(\text{program score}\mid \text{perturbation }g) -
\operatorname{mean}(\text{program score}\mid \text{WT}).
\]

Each source is supplied as a data frame with a `ko` column and one numeric
effect column named after that source. The `source_effects` input is a named
list of these data frames. The
[manuscript-code repository](https://github.com/yub-hutch/CausalGRN-manuscript-code)
provides the complete program-scoring and source-effect preparation scripts.

## Download and inspect the tutorial data

The file is hosted in the public
[CausalGRN tutorial-data project](https://osf.io/zu7dn/). The checksum prevents
the workflow from silently using a different file.

```{r download-data}
data_url <- "https://osf.io/download/6a66cab03059ac3f76135bfb/"
data_file <- file.path("data", "rpe1-tutorial-data.rds")
expected_md5 <- "aa57b50b5885cf21d713f8ce3bf5130e"

if (!file.exists(data_file)) {
  dir.create(dirname(data_file), recursive = TRUE, showWarnings = FALSE)
  download.file(
    url = data_url,
    destfile = data_file,
    mode = "wb"
  )
}

actual_md5 <- unname(tools::md5sum(data_file))
stopifnot(identical(actual_md5, expected_md5))

tutorial_data <- readRDS(data_file)
```

The object is a named list:

- `count`: prepared cell-by-node counts used for skeleton inference;
- `Y`: adjusted cell-by-node expression, including the program score;
- `group`: one perturbation label per cell, with `"WT"` denoting the reference
  cells;
- `program_genes`: member genes of the transcriptional program;
- `source_effects`: K562 and HCT116 effects on the RPE1-defined program score,
  used to model how program effects transfer across contexts; and
- `heldout_deg`: observed DEG labels reserved for evaluating the three
  prediction targets.

```{r inspect-data}
count <- tutorial_data$count
Y <- tutorial_data$Y
group <- tutorial_data$group
program_genes <- tutorial_data$program_genes
source_effects <- tutorial_data$source_effects
heldout_deg <- tutorial_data$heldout_deg

targets <- c("GPS1", "SRP72", "SRSF7")
program_name <- "PC1"
ncores <- 10L

stopifnot(
  identical(dim(count), dim(Y)),
  identical(rownames(count), names(group)),
  identical(rownames(Y), names(group)),
  identical(colnames(count), colnames(Y)),
  all(program_genes %in% colnames(Y)),
  all(targets %in% colnames(Y)),
  !any(group %in% targets),
  identical(rownames(heldout_deg), targets),
  identical(colnames(heldout_deg), setdiff(colnames(Y), program_name)),
  is.logical(heldout_deg),
  all(rowSums(heldout_deg) > 0)
)

data.frame(
  cells = nrow(Y),
  nodes = ncol(Y),
  perturbation_groups = length(unique(group)),
  WT_cells = sum(group == "WT"),
  program_genes = length(program_genes)
)
```

Network inference and model fitting use WT cells and perturbations other than
the three prediction targets. The observed RPE1 outcomes for these targets are
reserved for the evaluation in the final section.

# Infer the causal GRN

## Infer the graph skeleton

`infer_skeleton_with_gene_program()` tests marginal and order-one conditional
independence among the input nodes. The arguments used here are:

- `count = count`: raw counts used by the sparsity-aware conditional-
  independence tests;
- `Y = Y`: adjusted expression used to calculate correlations and partial
  correlations;
- `alpha = 0.05`: the significance threshold for the conditional-independence
  tests;
- `min_abspcor = 0.02`: the minimum absolute partial correlation required to
  retain an edge;
- `pname = program_name`: the name of the program node;
- `pgenes = program_genes`: the member genes that must be connected to the
  program node in the skeleton; and
- `ncores = ncores`: the number of CPU cores used in parallel calculations.

The function returns the skeleton as an `igraph` object.

```{r infer-skeleton}
skeleton <- infer_skeleton_with_gene_program(
  count = count,
  Y = Y,
  alpha = 0.05,
  min_abspcor = 0.02,
  pname = program_name,
  pgenes = program_genes,
  ncores = ncores
)

data.frame(
  nodes = igraph::vcount(skeleton),
  edge_entries = igraph::ecount(skeleton)
)
```

## Estimate observed perturbation effects

`calc_perturbation_effect()` compares each observed perturbation group with WT
for every node:

- `Y = Y` supplies the adjusted expression values;
- `group = group` identifies the WT and perturbed cells and is aligned with the
  rows of `Y`; and
- `ncores = ncores` controls parallel calculation.

The returned table reports the WT-to-perturbation mean difference (`diff`),
Cohen's d (`cd`), test results, and adjusted p-values for every
perturbation-node pair. `adj_pv` is the larger of the adjusted Wilcoxon and
t-test p-values and is used for graph orientation.

```{r estimate-effects}
perturbation_stat <- calc_perturbation_effect(
  Y = Y,
  group = group,
  ncores = ncores
)

head(
  perturbation_stat[, c("ko", "gene", "diff", "cd", "adj_pv")]
)
```

## Orient the skeleton

`infer_causalgrn_with_gene_program()` uses significant observed perturbation
effects to direct the skeleton:

- `graph = skeleton` supplies the graph to orient;
- `stat = perturbation_stat` supplies the observed effects;
- `alpha = 0.05` is the significance threshold applied to `stat$adj_pv`;
- `pname = program_name` and `pgenes = program_genes` identify the program node
  and its member genes; and
- `conservative = TRUE` retains both possible directions unless the
  perturbation evidence supports one direction.

The function returns an `igraph` object. A single `from`-to-`to` row is an
oriented edge; reciprocal rows indicate that the direction remains unresolved.

```{r orient-skeleton}
causal_graph <- infer_causalgrn_with_gene_program(
  graph = skeleton,
  stat = perturbation_stat,
  alpha = 0.05,
  pname = program_name,
  pgenes = program_genes,
  conservative = TRUE
)

data.frame(
  nodes = igraph::vcount(causal_graph),
  edge_entries = igraph::ecount(causal_graph)
)

edge_table <- igraph::as_data_frame(causal_graph, what = "edges")
edge_table <- edge_table[
  order(edge_table$from, edge_table$to),
  c("from", "to", "absPcorMin")
]
knitr::kable(head(edge_table, 10), digits = 3)
```

# Fit the perturbation-effect models

## Fit using reference source effects

`fit_expression_model_with_gene_program()` fits two linked model components:

1. one linear expression model per node, using its incoming GRN neighbors as
   predictors and cells in which that modeled node is unperturbed; and
2. portability regressions that relate source-context effects on the
   RPE1-defined program to its observed RPE1 effects across perturbations
   available during fitting.

Its arguments are:

- `Y = Y`, `group = group`, and `graph = causal_graph`: the RPE1 expression,
  perturbation labels, and inferred GRN;
- `source_effects = source_effects`: the prepared K562 and HCT116 effects on
  the RPE1-defined program score;
- `pname = program_name` and `pgenes = program_genes`: the program definition;
- `ncores = ncores`: the number of CPU cores used to fit node-level models; and
- `alpha = 0.1`: the F-test significance threshold for accepting a portability
  regression.

The result contains `B_matrix`, whose rows are predictors and columns are
modeled nodes, and `portability_models`, which are used to predict the RPE1
program-node effect for a new perturbation.

```{r fit-model}
expression_model <- fit_expression_model_with_gene_program(
  Y = Y,
  group = group,
  graph = causal_graph,
  source_effects = source_effects,
  pname = program_name,
  pgenes = program_genes,
  ncores = ncores,
  alpha = 0.1
)

data.frame(
  coefficient_rows = nrow(expression_model$B_matrix),
  modeled_nodes = ncol(expression_model$B_matrix)
)
```

## Fit when reference source effects are unavailable

When the target dataset contains an influential program but suitable reference
source effects are unavailable, `fit_expression_model()` can fit the same
program-containing graph directly from the target dataset. `Y = Y`,
`group = group`, and `graph = causal_graph` supply the adjusted expression,
perturbation labels, and inferred graph; `ncores = ncores` controls parallel
model fitting. The function returns the regulatory coefficient matrix.

```{r fit-model-without-source-effects}
expression_model_without_source <- fit_expression_model(
  Y = Y,
  group = group,
  graph = causal_graph,
  ncores = ncores
)

dim(expression_model_without_source)
```

# Predict effects at specified intervention expression levels

## Specify the intervention levels

An **intervention expression level** is the value imposed on a target gene for
prediction: "What expression changes would follow if this gene were set to this
level?" Users choose this input before observing the downstream response.

For illustration, each target is set to the first percentile of its expression
among the RPE1 WT cells. Users can replace these values with other experimentally
meaningful levels.

```{r intervention-levels}
wt_expressions <- colMeans(Y[group == "WT", , drop = FALSE])

intervention_expressions <- vapply(
  targets,
  function(target) {
    unname(stats::quantile(
      Y[group == "WT", target],
      probs = 0.01
    ))
  },
  FUN.VALUE = numeric(1)
)

knitr::kable(
  data.frame(
    target = targets,
    WT_mean = unname(wt_expressions[targets]),
    intervention_expression_level = unname(intervention_expressions)
  ),
  digits = 3
)
```

## Predict using reference source effects

`predict_perturbation_effect_with_gene_program()` first converts each specified
target level to a change from its WT mean, predicts the accompanying RPE1
program-node change from the reference source effects, and propagates those
known changes through the fitted RPE1 model.

- `B = expression_model$B_matrix` supplies the fitted regulatory coefficients;
- `portability_models = expression_model$portability_models` supplies the
  fitted cross-context program-effect models;
- `source_effects = source_effects` supplies the K562 and HCT116 program effects
  for the prediction targets;
- `ko_expressions = intervention_expressions` supplies the specified target
  expression levels;
- `wt_expressions = wt_expressions` supplies the RPE1 WT reference means; and
- `pname = program_name` and `pgenes = program_genes` supply the program
  definition.

The returned matrix contains predicted expression changes from WT, with one
target per row and one modeled node per column.

```{r predict-effects}
predicted_effects <- predict_perturbation_effect_with_gene_program(
  B = expression_model$B_matrix,
  portability_models = expression_model$portability_models,
  source_effects = source_effects,
  ko_expressions = intervention_expressions,
  wt_expressions = wt_expressions,
  pname = program_name,
  pgenes = program_genes
)

dim(predicted_effects)
```

## Predict without reference source effects

`predict_perturbation_effect()` uses the target-dataset coefficient matrix to
propagate each specified intervention through the inferred graph, including
the program node. `B = expression_model_without_source` supplies the fitted
coefficients, `ko_expressions = intervention_expressions` supplies the
specified target levels, and `wt_expressions = wt_expressions` supplies the WT
reference means.

```{r predict-without-source-effects}
predicted_effects_without_source <- predict_perturbation_effect(
  B = expression_model_without_source,
  ko_expressions = intervention_expressions,
  wt_expressions = wt_expressions
)

dim(predicted_effects_without_source)
```

## Inspect the strongest predicted effects

The figure reports the five strongest predicted decreases and increases among
individual downstream genes for each intervention target.

```{r strongest-effects, fig.width = 9.5, fig.height = 4.8, fig.cap = "Five strongest predicted decreases and increases for each intervention target."}
prediction_table <- as.data.frame(
  as.table(predicted_effects),
  responseName = "predicted_change"
)
names(prediction_table)[1:2] <- c("target", "gene")
prediction_table$target <- as.character(prediction_table$target)
prediction_table$gene <- as.character(prediction_table$gene)
prediction_table <- prediction_table[
  prediction_table$gene != program_name &
    prediction_table$gene != prediction_table$target,
]

strongest_decreases <- prediction_table |>
  group_by(target) |>
  slice_min(
    order_by = predicted_change,
    n = 5,
    with_ties = FALSE
  ) |>
  mutate(direction = "Decrease")

strongest_increases <- prediction_table |>
  group_by(target) |>
  slice_max(
    order_by = predicted_change,
    n = 5,
    with_ties = FALSE
  ) |>
  mutate(direction = "Increase")

strongest_effects <- bind_rows(
  strongest_decreases,
  strongest_increases
) |>
  ungroup() |>
  arrange(target, predicted_change)

strongest_effects$plot_label <- factor(
  paste(strongest_effects$target, strongest_effects$gene, sep = "___"),
  levels = unique(
    paste(strongest_effects$target, strongest_effects$gene, sep = "___")
  )
)

ggplot(
  strongest_effects,
  aes(
    x = predicted_change,
    y = plot_label,
    fill = direction
  )
) +
  geom_vline(xintercept = 0, color = "grey60", linewidth = 0.4) +
  geom_col(width = 0.75) +
  facet_wrap(~target, scales = "free_y") +
  scale_y_discrete(labels = function(x) sub(".*___", "", x)) +
  scale_fill_manual(
    values = c("Decrease" = "#377EB8", "Increase" = "#E41A1C")
  ) +
  labs(
    x = "Predicted expression change from WT",
    y = NULL,
    fill = NULL
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank()
  )
```

# Evaluate held-out DEG ranking

The full RPE1 dataset supplies observed differential-expression labels for
GPS1, SRP72, and SRSF7. A gene is labeled as a DEG when
`calc_perturbation_effect()` applied to the full dataset gives
`adj_pv < 0.05`. These labels are used here after model fitting.

For each target, genes are ranked by the absolute predicted expression change
from the source-aware model. Each curve shows the proportion of held-out DEGs
among the top \(N\) ranked genes. The dashed line gives the DEG prevalence
expected under random ranking. Each curve extends through the observed number
of DEGs for that target.

```{r evaluate-deg-ranking, fig.width = 9.5, fig.height = 3.8, fig.cap = "Precision of held-out DEG recovery as genes are ranked by absolute predicted expression change. Dashed lines show DEG prevalence."}
precision_curves <- do.call(
  rbind,
  lapply(targets, function(target) {
    genes <- setdiff(
      intersect(colnames(predicted_effects), colnames(heldout_deg)),
      target
    )
    ranked_genes <- genes[
      order(abs(predicted_effects[target, genes]), decreasing = TRUE)
    ]
    is_deg <- heldout_deg[target, ranked_genes]
    top_n <- seq_len(sum(is_deg))

    data.frame(
      target = target,
      top_n = top_n,
      precision = cumsum(is_deg)[top_n] / top_n,
      DEG_prevalence = mean(is_deg)
    )
  })
)

precision_baselines <- unique(
  precision_curves[, c("target", "DEG_prevalence")]
)

ggplot(
  precision_curves,
  aes(x = top_n, y = precision)
) +
  geom_hline(
    data = precision_baselines,
    aes(yintercept = DEG_prevalence),
    linetype = "dashed",
    color = "grey50"
  ) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~target, nrow = 1, scales = "free_x") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "Top N genes ranked by absolute predicted change",
    y = "Held-out DEG precision"
  ) +
  theme_bw()
```

The
[manuscript-code repository](https://github.com/yub-hutch/CausalGRN-manuscript-code)
provides the dataset-wide cross-validation analysis.
